\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-11-21 paper/main.tex
another round of JRM edits

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique orderings for each
participant. First, we choose a word uniformly and at random from the set of
words on the to-be-presented list. Second, we compute the distances between the
chosen word's feature(s) value(s) and the corresponding feature(s) value(s) of
all yet-to-be-presented words. Third, we convert these distances (between the
previously presented word's feature values, $a$, and each of the $W$
yet-to-be-presented candidate words' feature values, $b_{i \in 1...W}$) to
similarity scores:

@@

\begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b_i) = \frac{\mathrm{similarity}(a,
b)}{\sum_{j=1}^W \mathrm{similarity}(a, b_j)},
\end{equation}
where in the denominator, $b_j$ takes on the feature value of each of the $W$
to-be-presented words. The resulting set of normalized similarity scores sums
to 1.

@@

Following some of our prior work~\citep{FitzEtal23, HeusEtal18a, HeusEtal21,
MannEtal22}, we used low-dimensional embeddings to help visualize how
participants' memory fingerprints change across lists
(Figs.~\ref{fig:fingerprint-trajectories}A,~\fingerprintTrajectoryRandom A). To
compute a shared embedding space across participants and experimental
conditions, we concatenated the full set of across-participant average
fingerprints (for all lists and experimental conditions) to create a large
matrix with number-of-lists (16) by number-of-conditions (10, including the
adaptive condition) rows and seven columns (one for each feature clustering
score, plus an additional temporal clustering score column). We used principal
components analysis to project the seven-dimensional observations into a
two-dimensional space (using the two principal components that explained the
most variance in the data). For two visualizations
(Figs.~\ref{fig:fingerprint-trajectories}B,~\fingerprintTrajectoryRandom B), we
computed an additional set of two-dimensional embeddings for the
\textit{average} fingerprints across lists within a given list grouping (i.e.,
early or late). For those visualizations, we averaged across the rows (for each
condition and group of lists) in the combined fingerprint matrix prior to
projecting it into the shared two-dimensional space. This yielded a single
two-dimensional coordinate for each \textit{list group} (in each condition),
rather than for each individual list. We used these embeddings solely for
visualization; all statistical tests were carried out in the original
(seven-dimensional) feature spaces.

@@

For a given list of words, if the values along two feature dimensions (e.g.,
category and size) are correlated, then the clustering scores for those two
dimensions will also be correlated. When lists are sorted along a given feature
dimension, the sorted feature values will also tend to be correlated with the
serial positions of the words in the list. This means that the temporal
clustering score will \textit{also} tend to be correlated with the clustering
scores for the sorted feature dimension. These correlations mean that it can be
difficult to specifically identify when participants are using one feature
versus another (or a manipulated feature versus temporal information) to
organize or search their memories.

@@

We used the distances between feature values for successively presented words
(see \textit{Defining feature-based distances}) to estimate ``event
boundaries'' where the feature values changed more than
usual~\citep{EzzyDava11, MannEtal16,RadvCope06, SwalEtal09, SwalEtal11,
DuBrDava16}. For each list, for each feature dimension, we computed the
distribution of distances between the feature values for successively presented
words. We defined event boundaries (e.g., Fig.~\ref{fig:recall-dynamics}B) as
occurring between any successive pair of words whose distances along the given
feature dimension were greater than one standard deviation above the mean for
that list. Note that, because event boundaries are defined for each feature
dimension, each individual list may contain several sets of event boundaries,
each at different moments in the presentation sequence (depending on the
feature dimension of interest).

@@

While holding the set of words (and the assignments of words to lists)
constant, we manipulated two aspects of participants' experiences of studying
each list. We sought to understand the effects of these manipulations on
participants' memories for the studied words. First, we added two additional
sources of visual variation to the individual word presentations: font color
and onscreen location. Importantly, these visual features were independent of
the meaning or semantic content of the words (e.g., word category, size of the
referent, etc.) and of the lexicographic properties of the words (e.g., word
length, first letter, etc.). We wondered whether this additional
word-independent information might facilitate recall~\citep[e.g., by providing
new or richer potential ways of organizing or retrieving memories of the
studied words;][]{HargEtal12, Mada21, MeinEtal20, DavaEtal03, SlamBarl79,
DrewMurd80, SochEtal09} or impair recall~\citep[e.g., by distracting or
confusing participants with irrelevant information][]{MarsEtal15, MarsEtal12,
Lang05, ReinEtal92}. Second, we manipulated the orders in which words were
studied (and how those orderings changed over time). We wondered whether
presenting the same list of words with different appearances (e.g., by
manipulating font size and onscreen location) or in different orders (e.g.,
sorted along one feature dimension versus another) might serve to influence how
participants organized their memories of the words~\citep[e.g., ][]{PolyKaha08,
MannEtal15}. We also wondered whether some order manipulations might be
temporally ``sticky'' by influencing how \textit{future} lists were
remembered~\citep[e.g., ][]{SiroEtal05, LohnEtal10, Whit27, Badd68,
DarlMurd71}.

@@

A core assumption of our approach is that each participant organizes their
memories in a unique way. We defined each participant's \textit{memory
fingerprint} as the set of their permutation-corrected clustering scores across
all dimensions we tracked in our study, including their six feature-based
clustering scores (category, size, length, first letter, color, and location)
and their temporal clustering score. Conceptually, a participant's memory
fingerprint describes their tendency to order, in their recall sequences (and
presumably, organize in memory), the studied words along each dimension. If
these memory fingerprints are truly unique to each participant, then we would
expect that the estimated fingerprints computed for a given participant, on
different lists, should be more similar than the estimated fingerprints
computed for different participants. We reasoned that the feature-rich
condition would provide the best opportunity to test this assumption, since the
clustering scores would not be potentially confounded by order manipulations.
To test our ``unique memory fingerprint'' assumption, we compared the
similarity (correlation) between the fingerprint from a single list (from one
participant) and (a) the average fingerprint from all other lists from the same
participant versus (b) the average fingerprint from each other participant
(across all of their lists). Repeating this procedure for all lists and
participants, we found that participants' fingerprints for a held-out list were
reliably more similar to their fingerprints for other lists than they were to
other participants' fingerprints ($t(70280) = 5.077,~p < 0.001,~d =
0.162,~\mathrm{CI} = [3.086,~6.895]$). That within-participant fingerprint
similarity was greater than between-participant fingerprint similarity suggests
that participants' memory fingerprints are relatively stable across lists, and
that each participant's fingerprint is unique to them.

@@

participants in the feature-rich condition. Participants in the Participants in
the color-ordered condition also showed a trending increase in memory
performance on early lists. Participants' performances on early lists in all of
the other order manipulation conditions were indistinguishable from performance
on the early feature-rich lists. We also compared participants' temporal
clustering on early lists in each of these conditions to their temporal
clustering on early lists in the feature-rich condition

@@

both of the semantically ordered conditions exhibited stronger temporal
clustering on early lists (vs.~early feature-rich lists). Participants in the
length-ordered condition tended to exhibit \textit{less} temporal clustering on
early lists relative to early feature-rich lists, whereas participants in the
first letter-ordered condition exhibited stronger temporal clustering on early
lists. Participants in the visually ordered conditions showed similar temporal
clustering on early lists relative to early feature-rich lists. We also also
compared feature-based clustering on early lists across the order manipulation
and feature-rich conditions. Since these results were similar across both
semantic conditions (category and size;

@@

The above comparisons between memory performance on early lists in the order
manipulation and feature-rich conditions highlight how sorted lists are
remembered differently from random lists. We also wondered how sorting lists
along each feature dimension influenced memory relative to sorting lists along
the other feature dimensions (accuracy:
Tab.~\ref{tab:order-manip-vs-feature-rich-accuracy-early}; temporal clustering:
Tab.~\ref{tab:order-manip-vs-feature-rich-temporal-early}; feature-based
clustering: Tab~\ref{tab:feature-comparisons-early}). Participants trended
towards remembering early lists that were sorted semantically better than those
sorted lexicographically. Participants also remembered visually sorted lists
better than lexicographically sorted lists. However, participants showed no
reliable differences in recall for semantically versus visually sorted lists.
Participants temporally clustered semantically sorted lists more strongly than
lists sorted either lexicographically or visually, but did not show reliable
differences in temporal clustering on lexicographically versus visually sorted
lists. Participants also showed reliably more semantic clustering on
semantically sorted lists than lexicographically or visually sorted lists; more
lexicographic clustering on lexicographically sorted lists than semantically or
visually sorted lists; and more visual clustering on visually sorted lists than
semantically or lexicographically sorted lists. In summary, sorting lists by
different features appeared to have slightly different effects on overall
memory performance and temporal clustering. Participants also tended to cluster
their recalls along a given feature dimension more when the studied lists were
(versus were not) sorted along that dimension.

@@

conditions: $r(4) = 0.064,~p = 0.903,~\mathrm{CI} = [-0.972,$). Taken together,

@@

If participants show different sensitivities to order manipulations, how do
their behaviors carry over to later lists? We found that participants who
showed strong feature clustering on early lists often tended to show strong
feature clustering on late lists (Fig.~\ref{fig:clustering-carryover}A; overall
across participants and conditions: $r(179) = 0.591,~p < 0.001,~\mathrm{CI} =
[0.472,~0.682]$; category: $r(28) = 0.590,~p < 0.001,~\mathrm{CI} = [0.354,~
0.756]$; size: $r(28) = 0.488,~p = 0.006,~\mathrm{CI} = [0.134,~0.732]$;
length: $r(28) = 0.384,~p = 0.036,~\mathrm{CI} = [0.040,~0.681]$; first letter:
$r(28) = 0.202,~p = 0.284,~\mathrm{CI} = [-0.273,~0.620]$; color: $r(29) =
-0.183,~p = 0.325,~\mathrm{CI} = [-0.562,~0.258]$; location: $r(28) = 0.031, p
= 0.870,~\mathrm{CI} = [-0.240,~0.296]$; across conditions: $r(4) = 0.942,~p =
0.005,~\mathrm{CI} = [0.442,~1.000]$). Although participants tended to show
weaker feature clustering on late lists
(Fig.~\ref{fig:fingerprint-trajectories}) on \textit{average}, the associations
between early and late lists for individual participants suggests that some
influence of early order manipulations may linger on late lists. We found that
participants who exhibited larger carryover in feature clustering (i.e.,
continued to show strong feature clustering on late lists) for the semantic
order manipulations (but not other manipulations) also tended to show a smaller
decrease in recall on early versus late lists
(Fig.~\ref{fig:clustering-carryover}B; overall: $r(179) = 0.307,~p < 0.001,
\mathrm{CI} = [0.148,~0.469]$; category: $r(28) = 0.350,~p = 0.058,~\mathrm{CI}
= [0.050,~0.642]$; size: $r(28) = 0.708,~p < 0.001,~\mathrm{CI} = [0.472,~
0.862]$; length: $r(28) = 0.205,~p = 0.276,~\mathrm{CI} = [-0.109,~0.492]$;
first letter: $r(28) = 0.081,~p = 0.672,~\mathrm{CI} = [-0.433,~0.597]$; color:
$r(29) = 0.155,~p = 0.406,~\mathrm{CI} = [-0.174,~0.541]$; location: $r(28) =
0.052,~p = 0.787,~\mathrm{CI} = [-0.307,~0.360]$; across conditions: $r(4) =
0.635,~p = 0.176,~\mathrm{CI} = [-0.924,~0.981]$. Participants who exhibited
larger carryover in feature clustering also tended to show stronger temporal
clustering on late lists (relative to early lists) for all but the category
condition (Fig.~\ref{fig:clustering-carryover}C; overall: $r(179) = 0.426,~p <
0.001,~\mathrm{CI} = [0.285,~0.544]$; category: $r(28) = 0.110,~p = 0.564,~
\mathrm{CI} = [-0.284,~0.442]$; all non-category conditions: all $r$s $\geq
0.406$, all $p$s $\leq 0.023$; across conditions: $r(4) = 0.649,~p = 0.163,
\mathrm{CI} = [-0.856, 0.988]$).

@@

We suggest two potential interpretations of these findings. First, it is
possible that some participants are more ``malleable'' or ``adaptable'' with
respect to how they organize incoming information. When presented with list of
items sorted along \textit{any} feature dimension, they will simply adopt that
feature as a dominant dimension for organizing those items and subsequent
(randomly ordered) items. This flexibility in memory organization might afford
such participants a memory advantage, explaining their strong recall
performance. An alternative interpretation is that each participant comes into
our study with a ``preferred'' way of organizing incoming information. If they
happen to be assigned to an order manipulation condition that matches their
preferences, then they will appear to be ``sensitive'' to the order
manipulation and also exhibit a high degree of carryover in feature clustering
from early to late lists. These participants might demonstrate strong recall
performance not because of their inherently superior memory abilities, but
rather because the specific condition they were assigned to happened to be
especially easy for them, given their pre-experimental tendencies. To help
distinguish between these interpretations, we designed an \textit{adaptive}
experimental condition (see \textit{Adaptive condition}). The primary
manipulation in the adaptive condition is that participants each experience
three key types of lists. On \textit{random} lists, words are ordered randomly
(as in the feature-rich condition). On \textit{stabilize} lists, the
presentation order is adjusted to be maximally similar to the current estimate
of the participant's memory fingerprint (see \textit{Online “fingerprint”
analysis}). Third, on \textit{destabilize} lists, the presentation order is
adjusted to be \textit{minimally} similar to the current estimate of the
participant's memory fingerprint (see \textit{Ordering ``stabilize'' and
``destabilize'' lists by an estimated fingerprint}). The orders in which
participants experienced each type of list were counterbalanced across
participants to help reduce the influence of potential list-order effects.
Because the presentation orders on stabilize and destabilize lists are adjusted
to best match each participant's (potentially unique) memory fingerprint, the
adaptive condition removes uncertainty about whether participants' assigned
conditions might just ``happen'' to match their preferred ways of organizing
their memories.

@@

We note that the multivariate aspect of the adaptive condition (i.e., sorting
lists simultaneously along multiple feature dimensions) provides an important
contrast with the order order manipulation conditions, where we sort lists
along only a single feature dimension in each condition. We found that
participants ``naturally'' clustered their recalls along multiple feature
dimensions, even when the lists they studied were not sorted along those
dimensions (as in the feature-rich condition). A caveat is that the
\textit{specific} feature dimensions participants tended to cluster along
varied across participants. One way to quantify the multidimensional nature of
participants' clustering tendencies is to sort each partipant's clustering
scores (for each of the six feature dimensions, along with a seventh dimension
to capture temporal clustering). We can then ask whether the distribution of
clustering scores at each ``rank'' within the sorted set of scores for each
participant has a mean that is reliably different from a chance value of 0.5.
We carried out these tests for each set of ranked scores, and found that
participants in the feature-rich condition reliably cluster their recalls along
at least three dimensions, including temporal clustering (which was often
ranked highest; Tab.~\ref{tab:ranked-clustering}).

@@

Several prior studies have examined how varying the richness or experiences, or
the extensive of encoding, can affect memory. Although specific details
differ~\citep{BoniEtal22}, in general these studies have found that richer and
more deeply or extensively encoded experiences are remembered
better~\citep{HargEtal12, Mada21, MeinEtal20}. Our findings help to elucidate
an additional factor that may contribute to these phenomenon. For example, our
finding that participants better remember ``feature-rich'' lists (where words'
appearances are varied) than ``reduced'' lists (where words' appearances are
held constant) only when those feature-rich lists are presented \textit{after}
reduced lists suggests that some factors that influence the richness or depth
of encoding may be relative, rather than absolute. In other words,
\textit{increases} in richness (e.g., relative to a recency-weighted baseline)
may be more important than the overall complexity or numbers of features.

Some prior studies have suggested that people can ``cue'' their memories using
different ``strategies'' or ``pathways'' for searching for the target
information. For example, modern accounts of free recall typically posit that
memory search typically begins by matching the current state of mental context
with the contexts associated with other items in memory~\citep{Kaha20}. Since
context is the defining hallmark of episodic memory~\citep{Tulv83},
context-based search can be described as an ``episodic'' pathway to recall.
When episodic cueing fails to elicit a match, participants may then search for
items that are similar to the current mental context or mental state along
other dimensions, such as semantic similarity~\citep{DavaEtal03, SochEtal09}.
These multiple pathways accounts of memory search also provide a potential
explanation of why participants might have an easier time remembering richer
stimuli (or experiences): richer stimuli and experiences might have more
features that could be used to cue memory search. Our work suggests that there
may be some additional factors at play with respect to the \textit{dynamics} of
these processes. In particular, we only observed memory benefits for ``richer''
stimuli when they were encountered after more ``impoverished'' stimuli (in the
reduced (early) condition). This suggests that the pathways available to recall
a given item may also depend on recent prior experiences.

>>>>

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique orderings for each
participant. First, we choose a word uniformly and at random from the set of
words on the to-be-presented list. Second, we compute the distances between the
chosen word's feature value(s) and the corresponding feature value(s) of all
yet-to-be-presented words. Third, we convert these distances (between the
previously presented word's feature values, $a$, and each of the $W$
yet-to-be-presented candidate words' feature values, $b_{i \in 1...W}$) to
similarity scores:

@@

\begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b_i) = \frac{\mathrm{similarity}(a,
b_i)}{\sum_{j=1}^W \mathrm{similarity}(a, b_j)},
\end{equation}
where in the denominator, $b_j$ takes on the feature value of each of the $W$
to-be-presented words. The resulting set of normalized similarity scores sums
to 1.

@@

Following some of our prior work~\citep{FitzEtal23, HeusEtal18a, HeusEtal21,
MannEtal22}, we used low-dimensional embeddings to help visualize how
participants' memory fingerprints change across lists
(Figs.~\ref{fig:fingerprint-trajectories}A,~\fingerprintTrajectoryRandom A). To
compute a shared embedding space across participants and experimental
conditions, we concatenated the full set of across-participant average
fingerprints (for all lists and experimental conditions) to create a large
matrix with number-of-lists (16) $\times$ number-of-conditions (10, excluding
the adaptive condition) rows and seven columns (one for each feature clustering
score, plus an additional temporal clustering score column). We used principal
components analysis to project the seven-dimensional observations into a
two-dimensional space (using the two principal components that explained the
most variance in the data). For two visualizations
(Figs.~\ref{fig:fingerprint-trajectories}B,~\fingerprintTrajectoryRandom B), we
computed an additional set of two-dimensional embeddings for the
\textit{average} fingerprints across lists within a given list grouping (i.e.,
early or late). For those visualizations, we averaged across the rows (for each
condition and group of lists) in the combined fingerprint matrix prior to
projecting it into the shared two-dimensional space. This yielded a single
two-dimensional coordinate for each \textit{list group} (in each condition),
rather than for each individual list. We used these embeddings solely for
visualization; all statistical tests were carried out in the original
(seven-dimensional) feature spaces.

@@

For a given list of words, if the values along two feature dimensions (e.g.,
category and size) are correlated, then the clustering scores for those two
dimensions will also be correlated (Figs.~\clusterCorrs,~\featureCorrs). When
lists are sorted along a given feature dimension, the sorted feature values
will also tend to be correlated with the serial positions of the words in the
list. This means that the temporal clustering score will \textit{also} tend to
be correlated with the clustering scores for the sorted feature dimension.
These correlations mean that it can be difficult to specifically identify when
participants are using one feature versus another (or a manipulated feature
versus temporal information) to organize or search their memories.

@@

We used the distances between feature values for successively presented words
(see \textit{Defining feature-based distances}) to estimate ``event
boundaries'' where the feature values changed more than
usual~\citep{EzzyDava11, MannEtal16,RadvCope06, SwalEtal09, SwalEtal11,
DuBrDava16}. For each list, for each feature dimension, we computed the
distribution of distances between the feature values for successively presented
words. We defined event boundaries (e.g., Fig.~\ref{fig:recall-dynamics}B) as
occurring between any successive pair of words whose distances along the given
feature dimension were greater than two standard deviations above the mean for
that list. Note that, because event boundaries are defined for each feature
dimension, each individual list may contain several sets of event boundaries,
each at different moments in the presentation sequence (depending on the
feature dimension of interest).

@@

While holding the set of words (and the assignments of words to lists)
constant, we manipulated two aspects of participants' experiences of studying
each list. We sought to understand the effects of these manipulations on
participants' memories for the studied words. First, we added two additional
sources of visual variation to the individual word presentations: font color
and onscreen location. Importantly, these visual features were independent of
the meaning or semantic content of the words (e.g., word category, size of the
referent, etc.) and of the lexicographic properties of the words (e.g., word
length, first letter, etc.). We wondered whether this additional
word-independent information might facilitate recall~\citep[e.g., by providing
new or richer potential ways of organizing or retrieving memories of the
studied words;][]{HargEtal12, Mada21, MeinEtal20, DavaEtal03, SlamBarl79,
DrewMurd80, SochEtal09} or impair recall~\citep[e.g., by distracting or
confusing participants with irrelevant information;][]{MarsEtal15, MarsEtal12,
Lang05, ReinEtal92}. Second, we manipulated the orders in which words were
studied (and how those orderings changed over time). We wondered whether
presenting the same list of words with different appearances (e.g., by
manipulating font size and onscreen location) or in different orders (e.g.,
sorted along one feature dimension versus another) might serve to influence how
participants organized their memories of the words~\citep[e.g., ][]{PolyKaha08,
MannEtal15}. We also wondered whether some order manipulations might be
temporally ``sticky'' by influencing how \textit{future} lists were
remembered~\citep[e.g., ][]{SiroEtal05, LohnEtal10, Whit27, Badd68,
DarlMurd71}.

@@

A core assumption of our approach is that each participant organizes their
memories in a unique way. We defined each participant's \textit{memory
fingerprint} as the set of their permutation-corrected clustering scores across
all dimensions we tracked in our study, including their six feature-based
clustering scores (category, size, length, first letter, color, and location)
and their temporal clustering score. Conceptually, a participant's memory
fingerprint describes their tendency to order, in their recall sequences (and
presumably, organize in memory), the studied words along each dimension. If
these memory fingerprints are truly unique to each participant, then we would
expect that the estimated fingerprints computed for a given participant, on
different lists, should be more similar than the estimated fingerprints
computed for different participants. We reasoned that the feature-rich
condition would provide the best opportunity to test this assumption, since the
clustering scores would not be potentially confounded by order manipulations.
To test our ``unique memory fingerprint'' assumption, we compared the
similarity (correlation) between the fingerprint from a single list (from one
participant) and (a) the average fingerprint from all other lists from the same
participant versus (b) the average fingerprint from each other participant
(across all of their lists). Repeating this procedure for all lists and
participants, we found that participants' fingerprints for a held-out list were
reliably more similar to their fingerprints for other lists than they were to
other participants' fingerprints ($t(70280) = 5.077,~p < 0.001,~d =
0.162,~\mathrm{CI} = [3.086,~6.895]$). That within-participant fingerprint
similarity (across lists) was greater than across-participant fingerprint
similarity suggests that participants' memory fingerprints are relatively
stable across lists, and that each participant's fingerprint is unique to them.

@@

participants in the feature-rich condition. Participants' performances on early
lists in all of the other order manipulation conditions were indistinguishable
from performance on the early feature-rich lists. We also compared
participants' temporal clustering on early lists in each of these conditions to
their temporal clustering on early lists in the feature-rich condition

@@

both of the semantically ordered conditions and the first letter-ordered
condition all exhibited stronger temporal clustering on early lists (vs.~early
feature-rich lists). Participants in the other order manipulation conditions
all showed similar temporal clustering on early lists relative to early
feature-rich lists. We also also compared feature-based clustering on early
lists across the order manipulation and feature-rich conditions. Since these
results were similar across both semantic conditions (category and size;

@@

The above comparisons between memory performance on early lists in the order
manipulation and feature-rich conditions highlight how sorted lists are
remembered differently from random lists. We also wondered how sorting lists
along each feature dimension influenced memory relative to sorting lists along
the other feature dimensions (accuracy:
Tab.~\ref{tab:order-accuracy-comparisons-early}; temporal clustering:
Tab.~\ref{tab:order-temporal-comparisons-early}; feature-based
clustering: Tab~\ref{tab:feature-comparisons-early}). Participants trended
towards remembering early lists that were sorted visually
better than lexicographically sorted lists. However, participants showed no
reliable differences in recall for semantically versus lexicographically or visually sorted lists.
Participants temporally clustered semantically sorted lists more strongly than
lists sorted either lexicographically or visually, but did not show reliable
differences in temporal clustering on lexicographically versus visually sorted
lists. Participants also showed reliably more semantic clustering on
semantically sorted lists than lexicographically or visually sorted lists; more
lexicographic clustering on lexicographically sorted lists than semantically or
visually sorted lists; and more visual clustering on visually sorted lists than
semantically or lexicographically sorted lists. In summary, sorting lists by
different features appeared to have slightly different effects on overall
memory performance and temporal clustering. Participants also tended to cluster
their recalls along a given feature dimension more when the studied lists were
(versus were not) sorted along that dimension.

@@

conditions: $r(4) = 0.064,~p = 0.903,~\mathrm{CI} = [-0.972,~0.965$). Taken together,

@@

If participants show different sensitivities to order manipulations, how do
their behaviors carry over to later lists? We found that participants who
showed strong feature clustering on early lists often tended to show strong
feature clustering on late lists (Fig.~\ref{fig:clustering-carryover}A; overall
across participants and conditions: $r(179) = 0.591,~p < 0.001,~\mathrm{CI} =
[0.472,~0.682]$; category: $r(28) = 0.590,~p < 0.001,~\mathrm{CI} = [0.354,~
0.756]$; size: $r(28) = 0.488,~p = 0.006,~\mathrm{CI} = [0.134,~0.732]$;
length: $r(28) = 0.384,~p = 0.036,~\mathrm{CI} = [0.040,~0.681]$; first letter:
$r(28) = 0.202,~p = 0.284,~\mathrm{CI} = [-0.273,~0.620]$; color: $r(29) =
-0.183,~p = 0.325,~\mathrm{CI} = [-0.562,~0.258]$; location: $r(28) = 0.031, p
= 0.870,~\mathrm{CI} = [-0.240,~0.296]$; across conditions: $r(4) = 0.942,~p =
0.005,~\mathrm{CI} = [0.442,~1.000]$). Although participants tended to show
weaker feature clustering on late lists
(Figs.~\ref{fig:fingerprints},~\ref{fig:barplots}) on \textit{average}, the
associations between early and late lists for individual participants suggests
that some influence of early order manipulations may linger on late lists. We
found that participants who exhibited larger carryover in feature clustering
(i.e., continued to show strong feature clustering on late lists) for the
semantic order manipulations (but not other manipulations) also tended to show
a smaller decrease in recall on early versus late lists
(Fig.~\ref{fig:clustering-carryover}B; overall: $r(179) = 0.307,~p < 0.001,
\mathrm{CI} = [0.148,~0.469]$; category: $r(28) = 0.350,~p = 0.058,~\mathrm{CI}
= [0.050,~0.642]$; size: $r(28) = 0.708,~p < 0.001,~\mathrm{CI} = [0.472,~
0.862]$; length: $r(28) = 0.205,~p = 0.276,~\mathrm{CI} = [-0.109,~0.492]$;
first letter: $r(28) = 0.081,~p = 0.672,~\mathrm{CI} = [-0.433,~0.597]$; color:
$r(29) = 0.155,~p = 0.406,~\mathrm{CI} = [-0.174,~0.541]$; location: $r(28) =
0.052,~p = 0.787,~\mathrm{CI} = [-0.307,~0.360]$; across conditions: $r(4) =
0.635,~p = 0.176,~\mathrm{CI} = [-0.924,~0.981]$. Participants who exhibited
larger carryover in feature clustering also tended to show stronger temporal
clustering on late lists (relative to early lists) for all but the category
condition (Fig.~\ref{fig:clustering-carryover}C; overall: $r(179) = 0.426,~p <
0.001,~\mathrm{CI} = [0.285,~0.544]$; category: $r(28) = 0.110,~p = 0.564,~
\mathrm{CI} = [-0.284,~0.442]$; all non-category conditions: all $r$s $\geq
0.406$, all $p$s $\leq 0.023$; across conditions: $r(4) = 0.649,~p = 0.163,
\mathrm{CI} = [-0.856, 0.988]$).

@@

We suggest two potential interpretations of these findings. First, it is
possible that some participants are more ``malleable'' or ``adaptable'' with
respect to how they organize incoming information. When presented with list of
items sorted along \textit{any} feature dimension, they will simply adopt that
feature as a dominant dimension for organizing those items and subsequent
(randomly ordered) items. This flexibility in memory organization might afford
such participants a memory advantage, explaining their strong recall
performance. An alternative interpretation is that each participant comes into
our study with a ``preferred'' way of organizing incoming information. If they
happen to be assigned to an order manipulation condition that matches their
preferences, then they will appear to be ``sensitive'' to the order
manipulation and also exhibit a high degree of carryover in feature clustering
from early to late lists. These participants might demonstrate strong recall
performance not because of their inherently superior memory abilities, but
rather because the specific condition they were assigned to happened to be
especially easy for them, given their pre-experimental tendencies. To help
distinguish between these interpretations, we designed an \textit{adaptive}
experimental condition (see \textit{Adaptive condition}). The primary
manipulation in the adaptive condition is that participants each experience
three key types of lists. On \textit{random} lists, words are ordered randomly
(as in the feature-rich condition). On \textit{stabilize} lists, the
presentation order is adjusted to be maximally similar to the current estimate
of the participant's memory fingerprint (see \textit{Online “fingerprint”
analysis}). Third, on \textit{destabilize} lists, the presentation order is
adjusted to be \textit{minimally} similar to the current estimate of the
participant's memory fingerprint (see \textit{Ordering ``stabilize'' and
``destabilize'' lists by an estimated fingerprint}). The orders in which
participants experienced each type of list were counterbalanced across
participants to help reduce the influence of potential list-order effects.
Because the presentation orders on stabilize and destabilize lists are adjusted
to best match each participant's personal memory fingerprint, the adaptive
condition removes uncertainty about whether participants' assigned conditions
might just ``happen'' to match their preferred ways of organizing their
memories.

@@

We note that the multivariate aspect of the adaptive condition (i.e., sorting
lists simultaneously along multiple feature dimensions) provides an important
contrast with the other order manipulation conditions, where we sort lists
along only a single feature dimension in each condition. We found that
participants ``naturally'' clustered their recalls along multiple feature
dimensions, even when the lists they studied were not sorted along those
dimensions (as in the feature-rich condition). A caveat is that the
\textit{specific} feature dimensions participants tended to cluster along
varied across participants. One way to quantify the multidimensional nature of
participants' clustering tendencies is to sort each partipant's clustering
scores (for each of the six feature dimensions, along with a seventh dimension
to capture temporal clustering). We can then ask whether the distribution of
clustering scores at each ``rank'' within the sorted set of scores for each
participant has a mean that is reliably different from a chance value of 0.5.
We carried out these tests for each set of ranked scores, and found that
participants in the feature-rich condition reliably cluster their recalls along
at least three dimensions, including temporal clustering (which was often
ranked highest; Tab.~\ref{tab:ranked-clustering}). That the clustering scores
ranked in the top three dimensions were reliably greater than chance suggests
that participants organize their memories along at least three feature
dimensions, even when the words are studied in a random order.

@@

Several prior studies have examined how varying the richness or experiences, or
the extensive of encoding, can affect memory. Although specific details
differ~\citep{BoniEtal22}, in general these studies have found that richer and
more deeply or extensively encoded experiences are remembered
better~\citep{HargEtal12, Mada21, MeinEtal20}. Our findings help to elucidate
an additional factor that may contribute to these phenomenon. For example, our
finding that participants better remember ``feature-rich'' lists (where words'
appearances are varied) than ``reduced'' lists (where words' appearances are
held constant) only when those feature-rich lists are presented \textit{after}
reduced lists suggests that some factors that influence the richness or depth
of encoding may be relative, rather than absolute. In other words,
\textit{increases} in richness (e.g., relative to a recency-weighted baseline)
may be more important than the absolute complexity or numbers of meaningful
features.

Some prior studies have suggested that people can ``cue'' their memories using
different ``strategies'' or ``pathways'' for searching for the target
information. For example, modern accounts of free recall often posit that
memory search typically begins by matching the current state of mental context
with the contexts associated with other items in memory~\citep{Kaha20}. Since
context is the defining hallmark of episodic memory~\citep{Tulv83},
context-based search can be described as an ``episodic'' pathway to recall.
When episodic cueing fails to elicit a match, participants may then search for
items that are similar to the current mental context or mental state along
other dimensions, such as semantic similarity~\citep{DavaEtal03, SochEtal09}.
These multiple pathways accounts of memory search also provide a potential
explanation of why participants might have an easier time remembering richer
stimuli (or experiences): richer stimuli and experiences might have more
features that could be used to cue memory search. Our work suggests that there
may be some additional factors at play with respect to the \textit{dynamics} of
these processes. In particular, we only observed memory benefits for ``richer''
stimuli when they were encountered after more ``impoverished'' stimuli (in the
reduced (early) condition). This suggests that the pathways available to recall
a given item may also depend on recent prior experiences.
